Policy Implications of Coupled Economic and Ecological Dynamics in Aquatic Systems

### 2022-12-10 03_marlin_paper.qmd
circilng towards something

---
title: "Policy Implications of Coupled Economic and Ecological Dynamics in Aquatic Systems"
author:
    - Daniel Ovando
    - Darcy Bradley
    - Echelle Burns
    - Lennon Thomas
    - James Thorson
date: "`r Sys.Date()`"
format:
  pdf:
    toc: true
    fig-pos: 'H'
  html: default
  docx: default
execute:
  echo: false
  message: false
  warning: false
bibliography: references.bib
params:
  run_name: ["v0.1"]
abstract: "Very rough cut. Communities around the world are increasingly looking to spatial strategies for managing marine ecosystems. However, the design of spatial management strategies faces numerous challenges. Marine ecosystems are highly dynamic and opaque, and extractive entities such as fishing fleets respond to changes in marine ecosystems in highly complex ways. We present a framework for efficient simulation of the bio-economic dynamics of marine systems, designed to support both management and research. We demonstrate the model's capabilities by focusing on two case studies on the conservation and food security impacts of marine protected areas (MPAs). We highlight how coupled bio-economic dynamics can result in complex trade-offs between conservation, fisheries yields, distributional outcomes in simulated coral reef and plegaic ecosystems. "
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
---

XX explain hyperallometry option

xx add in weird option 5 density dependence

@@

theme_set(theme_classic(base_size = 12, base_family = "Helvetica") + theme(strip.background = element_blank()))

@@

Communities around the world are increasingly looking to spatial
strategies for managing marine ecosystems. For example, the "30x30"
movement calls for implementation of marine protected areas, a form of
spatial management, across 30% of the world's oceans by 2030. While the
objectives of individual spatial management strategies vary widely,
spatial management strategies are often supported based on their ability
to provide protection to broad swaths of marine ecosystems with one
management action, with potentially fewer costs and complications than
for example assessing, setting, and enforcing catch quotas for
individual species.

However, the design of spatial management strategies faces numerous
challenges. Marine ecosystems are highly dynamic and opaque, and
extractive entities such as fishing fleets respond to changes in marine
ecosystems in highly complex ways. These dynamics post challenges even
for single-species assessment and management strategies [@field2006a],
which are only amplified when we consider management policies
specifically designed around spatial features in the oceans. For
example, policies such as MPAs designed for both conservation and food
security must consider factors such as the optimal size and placement of
a protected area network given a wide array of life history,
exploitation levels, fleet dynamics, and preferred habitats, all of
which may change over time, particularly given the impacts of climate
change.

These challenges are amplified by the fact that our knowledge regarding
the design and implementation of spatial management strategies such as
MPAs is largely based on theoretical studies; the size of protected
areas [@ovando2021a] and time required[@nickols2019] to see substantial
effects from many MPAs means that we have relatively little empirical
evidence as to their large-scale effects under a range of
social-ecological settings [@mcclanahan2021].

Many models have been developed to support the theory and design of
spatial management strategies (see @fulton2015 for a review of existing
methods). However, they have tended to fall into one of either highly
complex and tactical models designed for use in a specific location, or
to be highly stylized representations intended to provide theoretical
insights with less relevance for highly specific design questions. This
complex landscape presents a challenge to stakeholders charged with
marine spatial planning exercises lacking accessible models suited for
more tactical applications, and a barrier to the scientific community,
wherein comparisons across spatial management simulation studies are
clouded by discrepancies in underlying model structures beyond
differences in the core phenomenon in question.

In part to help address this challenge, we present here a novel
bio-economic modeling framework called `marlin` that allows for
efficient simulation of spatio-temporal biological and economic
dynamics. `marlin` allows users to simulation the impacts spatial
policies such as MPAs across a range of species targeted by various
fleets across heterogeneous and dynamic seascapes. Here we present the
core methods and functionality of this framework, and demonstrate its
use in three xx case-study applications. In the first, we show how
`marlin` can be used to assess critical drivers for positive outcomes
and unintended consequences of MPAs as a bycatch-mitigation strategy. In
the second, we present a more tactical application, demonstrating how
`marlin` can be used to compare and design MPA networks to achieve
societal objectives.

@@

`marlin` follows a "models of intermediate complexity" (MICE) philosophy
[@plagányi2014] targeted specifically towards capturing heterogeneous
habitats, movement dynamics, and technical interactions across species.
In its general form, `marlin` simulates the behavior of populations of
spatially explicit biologically independent age-structured animal
populations affected by one or more fishing fleets in time steps
specified by the user (generally annual or seasonal). While `marlin`
does not simulate species interactions, fishing fleets in `marlin` are
capable of targeting and affecting multiple species simultaneously. This
allows the model to simulate processes such as fisheries bycatch or
effort displacement in a way that accounts for the ways that changes in
fleet behavior might affect multiple species in the system. The ability
to track the impacts of fishing fleets across multiple species
simultaneously is particularly important as very few fishing fleets are
truly single species.

Below we provide a summary of the `marlin` model, as well as details of
the generalized simulation exercises and case study analyses. We note
that `marlin` is a complex package with many options to control detailed
behaviors. Given that these specific features will evolve over time, we
do not describe in detail every option of the model here. Rather, we
focus on the equations and methods describing the core foundations of
our framework for efficiently simulating bio-economic dynamics in
aquatic systems.

@@

@@

A fishing fleet is defined in `marlin` by a set of fishers that can have
different catchability, prices, and selectivity for individual species
within a simulation (each of which we denote as a *metier*). For
example, both a longline and purse seine fleet may capture bigeye
(*Thunnus obesus*) and skipjack (*Katswamus pelamus*) tunas, but the
longline fleet is much more likely to catch bigeye than skipjack, and
*vice versa*. Each fishing fleet (*f*) generates catches or yield (*Y*),
revenues (*R*), costs (*C*) from individual species (*s*) that it
targets in a given time step (*t*) and patch (*p*). The fleet then makes
decisions around how much and where to fish, subject to regulatory
constraints, based on th total profitability across all species in space
and time.

Revenues *R* for fleet *f* in time step *t* are a function of the total
amount caught of each species and its price ($\Pi$). The amount caught
is a function of the selectivity at age of each species for each fleet
($\alpha$), the fishing efficiency of the fleet for that that species
(*q*), the amount of fishing effort of the fleet in question in that
patch in a given time step (*E*), and the total fishing mortality at age
(including all other fleets) for that particular species in that patch
and time (*u*).

$$
 R_{t,p,f} = \sum_s^S \sum_a^A \Pi_{s,f}\frac{\alpha_{a,s,f}q_{s,f}E_{t,p,f}}{u_{t,p,s,a}} \times Y_{t,p,s,a}
$$ {#eq-revenue}

*Y* is total catches, which is calculated through the Baranov equation
[@baranov1918], which appropriately accounts for the total instantaneous
mortality (both fishing and natural) and divides the total amount of the
population biomass (*b*) killed between natural (*m*) and fishing (*u*)
sources, the amount of biomass killed through fishing being assigned to
yield *Y*.

@@

where *b* is biomass and *u* is total fishing mortality, given by

$$
u_{t,p,s,a} = \sum_{f=1}^F\alpha_{a,s,f}q_{s,f}E_{t,p,f}
$$ {#eq-u}

Selectivity $\alpha$ is modeled through either a logistic form or a
dome-shaped form. The logistic form is based on the lengths *l* at which
50% of individuals are selected by the fishing gear, $l^{sel}$, and
$\delta$ which is the difference between the length at 50% selectivity
and the length at 95% selectivity.

@@

$$
\alpha_{a,s,f}= \frac{1}{\sigma_{s,f} \sqrt{2\pi}}e^{-0.5 \left( \frac{l_{a,s} - l^{sel}}{\sigma_{s,f}}  \right)^2}
$$ Lastly *z* is total mortality, the sum of fishing mortality and
natural mortality (*m*).

@@

Costs *C* are calculated as a function of $\gamma_f$, that the scales
the cost per unit effort for fleet *f*,$\beta$ allows for the cost of
effort to scale non-linearly, and $\eta$ is an additional cost per unit
effort of fishing in each patch *p*, allowing for the model to account
for travel costs for different patches. Travel costs are calculated
based on the euclidean distance of each patch to the nearest port of a
given fleet, users can specify any number from zero to the number of
patches of port locations. When no ports are specified, travel costs are
zero.

$$C_{t,p,f} = \gamma_f  \left( E_{t,p,f}^{\beta_f} + \eta_{f,p} E_{t,p,f} \right)$$

@@

When MPAs are implemented, effort that used to operate inside the MPAs
can either leave the fishery, or be redistributed outside the the MPA
(the default behavior). The model can also accept total catch quotas for
each species in each time step. When quotas are activated, if the total
catch across all fleets for a given species under the model's effort
dynamics would exceed the allowable quota for that species, the total
effort for each fleet is decrease proportional to its contribution to
the total catch until the quota is satisfied. As an alternative to catch
quotas, users can input an effort cap per fleet, which prevents effort
from exceeding a given amount under open-access dynamics, though effort
may be reduced below this cap if required by the profit equation.
`marlin` also allows users to specify closed fishing seasons for one or
more species in the system. During closed seasons, no effort is
allocated to patches that would result in capture of the covered
species, potentially resulting in zero effort if the species cannot be
avoided. Under these conditions, the amount of effort in the next open
season is a function of the effort and profits in the last open fishing
season.

@@

Users can tune their fleets in one of two ways. First, they can specify
a target exploitation rate *u* for each species in their model run.
Taking all the other parameters of the model as given, `marlin` then
adjusts the catchability coefficients *q\_{f,s}* for each fleet *f* and
species *s* such that the equilibrium exploitation rate for each species
matches the desired level. Second, they can specify a target depletion,
defined as total spawning stock biomass under fishing divided by total
unfished spawning stock biomass, and the model will adjust the
catchability coefficients *q\_{f,s}* for each fleet *f* and species *s*
such that the equilibrium depletion for each species matches the desired
level.

@@

Each fishing fleet decides how to allocate its effort in space based on
one of four possible spatial allocation strategies. The ideal free
distribution (IFD) is the standard method for distributing fishing
fleets in spaces (see @gillis2003 and references therein). However, the
IFD presents a number of complications for our model. First, under the
constant effort model, the total amount of effort required to satisfy
the IFD may not equal the total amount of effort required by the model.
Second, even if we were to implement the IFD for a single fleet, the
model is capable of tracking multiple fleets at once. This means that
the IFD for a single fleet would be a game-theoretic outcome of the
actions of each of the individual fleet conditional on the actions of
all other fleets. While possible to implement this would slow down our
model runs to the point of practically preventing large-scale evaluation
of the effects of MPAs on bycatch.

@@

1.  Revenue per unit effort (RPUE): The fleet distributes itself in
    space based on revenue per unit effort
2.  Revenue: The fleet distributes itself in space based on revenue
3.  Profit per unit effort (PPUE): The fleet distributes itself in space
    based on profit per unit effort
4.  Profit: The fleet distributes itself in space based on profit

Revenue based spatial distribution is not likely to be very realistic .
However, due to the complexity of parameterizing cost functions fleet
dynamics are often evaluated based on yield or revenue alone, and so we
include those scenarios here to evaluate the potential implications of
this choice. The decision on whether to allocate the fleet based on
absolute or relative (per unit effort) is more complex. When effort
represents the actions of separate and individual fishing actors (e.g.
independent fishing vessels), a per-unit-effort strategy may be more
realistic, in which fishers distribute themselves based on the expected
catch of their individual efforts [@hilborn1987]. Conversely, a system
defined by a sole owner seeking to maximize total profits might be
better represented by a fleet model based on total profits

@@

The underlying population model used is an age structured single-species
model in the manner of @ovando2021a. The population model requires a
large number of parameters. However, if the user supplies either a
scientific or common name for the species in question (scientific
preferred), the model will supply default values for that species based
on the values reported in FishLife [@thorson2020]. All of these values
can be replaced with user-supplied values as needed though, the defaults
simply ensure that reasonable and internally consistent values for the
species in question are supplied as a starting point.

@@

where *BH* is the Beverton-Holt recruitment function, *ssb* is
spawning-stock-biomass. $l_a$ is the mean length at at age, $l_sel$ is
the length at which on average 50% of individuals are selected by the
fishery, and $\delta_{sel}$ are the additional units of length at which
on average 95% of fish are selected by the fishery.

@@

Where *u* is a log-normally distributed term allowing for variation in
the length at age relationship

$$
u_s \sim normal(0,\sigma_s)
$$ {#eq-sigma}

@@

$$
  ssb_{t,p,s} = \sum_{a=1}^Aw_{a,s,t}^{\gamma_s} mat_{t,a} n_{t,p,s,a}
$$ {#eq-ssb}

@@

Recruitment follows Beverton-Holt dynamics. We do however allow for
three variants in the timing of density dependence:

@@

where *hab* is a vector of recruitment habitat quality by patch degined
by

$$hab_p = r0_p / \sum_p^P{r0_p}$$

@@

where **d^l^** is the recruitment movement matrix (xx make this notation
consistent with the movement block)

@@

`marlin`'s movement dynamics are based on a continuous-time Markov chain
(CTMC), as described in @thorson2021a. Within this framework, the model
allows for movement to be broken down into three components of
*diffusion*, *advection* (follow currents), and *taxis* (active movement
towards say a preferred habitat). For now, our model focuses just on the
diffusion and taxis components of this model, though future extensions
could easily incorporate advection vectors from say ROMs models
(citations). In this way then, `marlin` allows users to run anything
from a simple Gaussian dispersal kernel up to a system governed by
species that passively diffuse around a dynamic thermal range.

@thorson2021a described the exact steps of the CTMC approach in detail.
We briefly the general CTMC method here, but see @thorson2021a for a
detailed description of the method. Under this framework, movement from
each patch to each other patch in the system in a given timestep *t* for
life stage *g* of species *s* is defined by a movement matrix
$\boldsymbol{M}_{t,s,g}$. $\boldsymbol{M}_{t,s,g}$ is calculated as the
matrix exponent of the sum of the diffusion and taxis matrices scaled by
the width of the time step that the model is running under.

$$
\boldsymbol{M}_{t,s,g} = e^{\Delta t(\boldsymbol{D}_{t,s,g} + \boldsymbol{X}_{t,s,g})} 
$$ {#eq-movement}

Where $\boldsymbol{D}$ is the diffusion matrix and $\boldsymbol{X}$ is
the taxis matrix.

The diffusion matrix is a function of a user specified diffusion rate,
see @thorson2021a. We slightly modify the taxis matrix in the following
manner

xx need to fix notation somewhere here, see this line of code for much
clearer depiction of what's happening, since sometimes code is cleaner.
https://github.com/DanOvando/marlin/blob/3563e7319ad0267f89789267a4c6ee58a9ad0d60/R/fish_class.R#L525
xx

$$
n_{t,p,s,a} = \begin{cases}
      = (1 + \lambda d_{s,f})max(0, h(p2_{t,s,g}) - h(p1_{t,s,g})) & \text{if p2 and p1 are adjacent}\\
     = -\sum_{g' \neq g1}\{ (1 + \lambda d_{s,f})max(0, h(p2_{t,s,g}) - h(p1_{t,s,g}))\} & \text{if p1 = p2}\\
     = 0 & \text{otherwise.}
  \end{cases}
$$ {#eq-taxis}

The $lambda d_{s,f}$ term is a scalar in which the magnitude of the
habitat gradient between patches can be scaled relative to the diffusion
parameter $d$.

For both the diffusion and taxis matrix, we allow for the inclusion of
physical barriers to movement (i.e. land). Pairs of patches that are
adjacent but in which one or both patches are a barrier to movement are
set as non-adjacent. The CTMC model then produces movement dynamics that
move around barriers rather than over them. This is different than
setting habitat in a patch to zero, which implies a patch that a given
animal will not move towards for its own sake, but can move over towards
another better patch.

While this CTMC approach to movement simulation is to date not commonly
seen in the marine modeling literature, it has numerous advantages that
warrant its seeming complexity. First, when only diffusion is present,
the model will generalize to the familiar dynamics of a Gaussian
dispersal kernel. However, the time component *t* in @eq-movement allows
the movement model to be run on arbitrarily fine time steps. Second, the
*taxis* model allow for more active habitat choices by species, allowing
us to simulate both preferences of species in space and time
efficiently.Lastly, this form has the advantage that the parameters are
directly estimable from real data. So, if provided with say CPUE data
and a tagging study, users can estimates the movement parameters in the
same manner as @thorson2021a, and then pass their estimated parameters
to `marlin` for simulation.

One challenge is that the coefficients of the model matrices inside
@eq-movement can be hard to interpret. To facilitate this, `marlin`
allows users to specify a maximum number of cells that adults or
recruits diffuse in an year. The model will then tune the diffusion
parameter and habitat gradients to achieve this desired level of
movement.

@@

### Use Cases

>>>>

---
title: "Policy Implications of Coupled Economic and Ecological Dynamics in Aquatic Systems"
author:
    - Daniel Ovando
    - Darcy Bradley
    - Echelle Burns
    - Lennon Thomas
    - James Thorson
date: "`r Sys.Date()`"
format:
  pdf:
    toc: true
    fig-pos: 'H'
  html: default
  docx: default
execute:
  echo: false
  message: false
  warning: false
  cache: false
bibliography: references.bib
params:
  run_name: ["v0.3"]
abstract: "Communities around the world are increasingly looking to spatial strategies for managing marine ecosystems. However, the design of spatial management strategies faces numerous challenges. Marine ecosystems are highly dynamic and opaque, and extractive entities such as fishing fleets respond endogenously to changes in marine ecosystems in ways that depend upon ecological and policy context. We present a framework for efficient simulation of the bio-economic dynamics of marine systems, designed to support both management and research. We demonstrate the model's capabilities by focusing on two case studies on the conservation and food security impacts of marine protected areas (MPAs). We highlight how coupled bio-economic dynamics can result in complex trade-offs between conservation, fisheries yields, and distributional outcomes. "
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
---

@@

theme_set(theme_classic(base_size = 10, base_family = "Helvetica") + theme(strip.background = element_blank()))

@@

Communities around the world are increasingly looking to spatial
strategies for managing marine ecosystems. For example, the "30x30"
movement calls for implementation of marine protected areas, a form of
spatial management, across 30% of the world's oceans by the year 2030.
While the objectives of individual spatial management strategies vary
widely, these types of policies are often supported based on their
ability to provide protection, along with potential broader benefits, to
swaths of marine ecosystems with one management action, with potentially
fewer costs and complications than for example assessing, setting, and
enforcing catch quotas for individual species.

However, the design of spatial management strategies faces numerous
challenges. Marine ecosystems are highly dynamic and opaque, and
extractive entities such as fishing fleets respond to changes in marine
ecosystems in highly complex ways. These dynamics pose challenges even
for single-species assessment and management strategies [@field2006a],
which are only amplified when we consider management policies
specifically designed around spatial features in the oceans. For
example, policies such as MPAs designed for both conservation and food
security must consider factors such as the optimal size and placement of
a protected area network given a wide array of life history, species
distributions, exploitation levels, fleet dynamics, and policy-dependent
behavior [@reimer2017], all of which may change over time, particularly
given the impacts of climate change.

These challenges are amplified by the fact that our knowledge regarding
the succesful design and implementation of spatial management strategies
such as MPAs is largely based on theoretical studies; the size
[@ovando2021a] and time required [@nickols2019] to see substantial
effects from many MPAs means that we have relatively little empirical
evidence as to their large-scale effects across a range of
social-ecological settings [@mcclanahan2021].

Many models have been developed to support the theory and design of
spatial management strategies (see @fulton2015 for a review of existing
methods). However, they have tended to fall into one of either highly
complex and tactical models designed for use in a specific location, or
to be highly stylized representations intended to provide theoretical
insights with less relevance for specific applications. This lack of
accessible models suited for more tactical application presents a
challenge to stakeholders charged with marine spatial planning
exercises. It also presents a barrier to the scientific community,
wherein comparisons across spatial management simulation studies are
clouded by discrepancies in underlying model structures beyond
differences in the core phenomenon in question.

To help address this challenge, we present a novel bio-economic modeling
framework called [`marlin`](https://danovando.github.io/marlin/) that
allows for efficient simulation of spatio-temporal biological and
economic dynamics. `marlin` allows users to simulate the impacts
fisheries management policies across a range of species targeted by
various fleets across heterogeneous and dynamic seascapes. This model
can help users seek Parteo-optimal solutions that produce win-wins or
minimize tradeoffs across multiple objectives, such as food security and
conservation [@lester2013].

Here we present the core methods and functionality of `marlin`, and
demonstrate its use in two case-study applications; a coral reef system
and a pelagic system. Our results show how interacting ecological,
economic, and design attributes can affect the degree of benefits,
costs, and tradeoffs between conservation and food security outcomes of
MPAs. More broadly, this paper both demonstrates the functionality of
our model and the critical need for considering sufficiently realistic
coupled ecological and economic dynamics in policy evaluations.

@@

`marlin` follows a "models of intermediate complexity" (MICE) philosophy
[@plagányi2014] targeted specifically towards capturing heterogeneous
habitats, movement dynamics, and the simultaneous impacts of fishing
fleets across multiple species (frequently called "technical
interactions"). In its general form, `marlin` simulates the behavior of
populations of spatially explicit biologically independent
age-structured animal populations affected by one or more fishing fleets
in time steps specified by the user (generally annual or seasonal).
While `marlin` does not simulate species interactions, fishing fleets in
`marlin` are capable of targeting and affecting multiple species
simultaneously. This allows the model to simulate processes such as
fisheries bycatch or effort displacement in a way that accounts for how
fleet behavior might affect multiple species in the system. The ability
to track the impacts of fishing fleets across multiple species
simultaneously is particularly important as very few fishing fleets are
truly single species.

As a demonstration of the use of this modeling framework, we explore two
general situations

1.  Trade-offs and distributional outcomes for food security and
    ecosystem biomass in a coral-reef fishery
2.  Implications of climate-driven range shifts for MPA design in a
    pelagic fishery

We chose these two examples to illustrate the use of the model in
contrasting systems in which MPAs are increasingly considered.

Below we provide a summary of the `marlin` model, as well as details of
the generalized simulation exercises and case study analyses. We note
that `marlin` is a complex software package with many options to control
detailed behaviors. Given that these specific features will evolve over
time, we do not describe in detail every option of the model here.
Rather, we focus on the equations and methods describing the core
foundations of our framework for efficiently simulating bio-economic
dynamics in aquatic systems.

@@

### Model Summary

`marlin` simulates the dynamics of one or more species, currently best
representing fish-like species, using age-structured population
dynamics. Each time step, species move throughout the simulated seascape
using both diffusion and "taxis" (active movement towards preferred
habitat), experiences natural and potentially fishing mortality, and if
in a spawning season spawn using one of the possible forms of density
dependence implemented in the model, with potentially auto-correlated
deviations from deterministic reproductive dynamics.

These species are caught by fishing fleets. A fishing fleet is defined
in `marlin` by a set of fishers that can have different fishing skill,
prices, and selectivity for individual species (each of which we denote
as a *metier*). For example, both a longline and purse seine fleet may
capture bigeye (*Thunnus obesus*) and skipjack (*Katswamus pelamus*)
tunas, but the longline fleet can be made much more likely to catch
bigeye than skipjack, and *vice versa*. Each fishing fleet distributes
its fishing effort in space according to a specified spatial allocation
rule, for example in proportion to profit-per-unit-effort, conditional
on the presence of any spatial restrictions such as MPAs.

Unconstrained by management, the total amount of effort exerted by each
fleet can follow one of two dynamics: open-access or constant-effort.
Under constant effort, the total amount of effort of each fleet is fixed
over time, with the possible exception of attrition due to MPA
placement. Under open-access, the total amount of effort in a given time
step is a function of the profitability of the fleet in the previous
time step, evolving until a bionomic equilibrium of zero total profits
is reached. Profitability is a function of the volume and price of each
species caught, as well as the cost of the total amount of fishing
effort per fleet and the distance from port that the fleet operates at.

The dynamics of each fleet can be modified by management in a variety of
ways. Managers can impose size limits for individual species within each
fleet. They can also impose total catch quotas for each species in each
time step. When quotas are activated, if the total catch across all
fleets for a given species under the model's effort dynamics would
exceed the allowable quota for that species, the total effort for each
fleet is decrease proportional to its contribution to the total catch
until the quota is satisfied. As an alternative to catch quotas,
managers can set an effort cap per fleet, which prevents effort from
exceeding a given amount under open-access dynamics, though effort may
be reduced below this cap if required by the profit equation (i.e. the
fleet can choose to fish less than the quota or effort cap). `marlin`
also allows users to specify closed fishing seasons for one or more
species in the system. During closed seasons, no effort is allocated to
patches that would result in capture of the covered species, potentially
resulting in zero effort if the species cannot be avoided. Under these
conditions, the amount of effort in the next open season is a function
of the effort and profits in the last open fishing season. Lastly,
managers can specify locations of no-take MPAs, which could in theory
change in size and location. When MPAs are implemented, fishing effort
that used to operate inside the MPAs can either leave the fishery, or be
redistributed outside the the MPA (the default behavior).

Users can use the model to explore general properties of spatially
explicit social-ecological systems, or can parameterize it to explore
key uncertainties of more tactical applications. As an example of the
later case, we provide two case studies here designed to reflect the
kinds of questions stakeholders might face if designing an MPA network
for a coral reef or pelagic system.

@@

Each fishing fleet (*f*) generates catches or yield (*Y*), revenues
(*R*), costs (*C*) from individual species (*s*) that it targets in a
given time step (*t*) and patch (*p*). The fleet then makes decisions
around how much and where to fish, subject to regulatory constraints,
based on th total profitability across all species in space and time.

Revenues *R* for fleet *f* in time step *t* are a function of the total
amount caught of each species (*Y*) and its price ($\Pi$). The amount
caught is a function of the selectivity at age of each species for each
fleet ($\alpha$), the fishing efficiency of the fleet for that that
species (also called "catchability", *q*), the amount of fishing effort
of the fleet in question in that patch in a given time step (*E*), and
the total fishing mortality at age (including all other fleets) for that
particular species in that patch and time (*u*).

$$
 R_{t,p,f} = \sum_s^{N_s} \sum_a^{N_a} \Pi_{s,f}\frac{\alpha_{a,s,f}q_{s,f}E_{t,p,f}}{u_{t,p,s,a}} \times Y_{t,p,s,a}
$$ {#eq-revenue}

Total catches or yield *Y* are calculated through the Baranov equation
[@baranov1918], which appropriately accounts for the total instantaneous
mortality (both fishing and natural) and divides the total amount of the
population biomass (*b*) killed between natural (*m*) and fishing (*u*)
sources, the amount of biomass killed through fishing being assigned to
yield *Y*.

@@

$$
u_{t,p,s,a} = \sum_{f=1}^{N_f}\alpha_{a,s,f}q_{s,f}E_{t,p,f}
$$ {#eq-u}

Selectivity at age $\alpha$ is modeled through either a logistic form or
a dome-shaped form. The logistic form is based on the lengths *l* at
which 50% of individuals are selected by the fishing gear, $l^{sel}$,
and $\delta$ which is the difference between the length at 50%
selectivity and the length at 95% selectivity.

@@

$$
\alpha_{a,s,f}= \frac{1}{\sigma_{s,f} \sqrt{2\pi}}e^{-0.5 \left( \frac{l_{a,s} - l^{sel}}{\sigma_{s,f}}  \right)^2}
$$ {#eq-dome}

Lastly *z* is total mortality, the sum of fishing mortality and natural
mortality (*m*).

@@

Costs *C* are calculated as a function of $\gamma_f$, that the scales
the cost per unit effort for fleet *f*,$\beta$ allows for the cost of
effort to scale non-linearly, and $\eta$ is an additional cost per unit
effort of fishing in each patch *p*, allowing for the model to account
for travel costs for different patches. Travel costs are calculated
based on the Euclidean distance of each patch to the nearest port of a
given fleet, users can specify any number from zero to the number of
patches of port locations. When no ports are specified, travel costs are
zero.

$$
C_{t,p,f} = \gamma_f  \left( E_{t,p,f}^{\beta_f} + \eta_{f,p} E_{t,p,f} \right)
$$ {#eq-travel-costs}

@@

This is essentially a Gompertz model for fishing effort, that has been
used for other theoretical studies of fishing effort dynamics
[@thorson2013].

@@

Users can tune their fleets in one of two ways. First, they can specify
a target exploitation rate *u* for each species in their model run.
Taking all the other parameters of the model as given, `marlin` then
adjusts the catchability coefficients $q_{f,s}$ for each fleet *f* and
species *s* such that the equilibrium exploitation rate for each species
matches the desired level. Second, they can specify a target depletion,
defined as total spawning stock biomass under fishing divided by total
unfished spawning stock biomass, and the model will adjust the
catchability coefficients $q_{f,s}$ for each fleet *f* and species *s*
such that the equilibrium depletion for each species matches the desired
level.

@@

Each fishing fleet decides how to allocate its effort in space based on
one of four possible spatial allocation strategies. The ideal free
distribution (IFD) is the standard method for distributing fishing
fleets in spaces (see @gillis2003 and references therein). However,
analytical solutions to the IFD present a number of complications for
our model. Primarily, our model is capable of tracking multiple fleets
at once. This means that the IFD for a single fleet would be a
game-theoretic outcome of the actions of each of the individual fleet
conditional on the actions of all other fleets. While possible to
implement this would slow down our model runs to the point of
practically preventing large-scale evaluation of the effects of MPAs on
bycatch.

@@

1.  Revenue per unit effort (RPUE): The fleet distributes itself in
    space based on the realized revenue per unit effort in each patch in
    the preceding time step
2.  Revenue: The fleet distributes itself in space based on the realized
    total revenue in each patch in the preceding time step
3.  Profit per unit effort (PPUE): The fleet distributes itself in space
    based on the realized profit per unit effort in each patch in the
    preceding time step
4.  Profit: The fleet distributes itself in space based on profit in
    each patch in the preceding time step

Revenue based spatial distribution is not likely to be very realistic;
in general we would expect fishing fleets to respond to profits on some
core level. However, due to the complexity of parameterizing cost
functions fleet dynamics are often evaluated based on yield or revenue
alone, and so we include those scenarios here to allow users to evalaute
the potential implications of this choice. The decision on whether to
allocate the fleet based on absolute or relative (per unit effort)
metrics is more complex. When effort represents the actions of separate
and individual fishing actors (e.g. independent fishing vessels), a
per-unit-effort strategy may be more realistic, in which fishers
distribute themselves based on the expected catch of their individual
efforts [@hilborn1987]. Conversely, a system defined by a sole owner
seeking to maximize total profits might be better represented by a fleet
model based on total profits

@@

The underlying population model used is an age structured single-species
model in the manner of @ovando2021a. The population model requires many
parameters. However, if the user supplies either a scientific or common
name for the species in question (scientific preferred), the model will
supply default values for that species based on the values reported in
FishLife [@thorson2020]. All of the life history values can be replaced
with user-supplied values as needed though, the defaults simply ensure
that reasonable and internally consistent values for the species in
question are supplied as a starting point.

`marlin`'s movement dynamics are based on a continuous-time Markov chain
(CTMC), as described in @thorson2021a. Within this framework, the model
allows for movement to be broken down into three components of
*diffusion*, *advection* (follow currents), and *taxis* (active movement
towards say a preferred habitat). For now, our model focuses just on the
diffusion and taxis components of this model, though future extensions
could easily incorporate advection vectors from oceanographic models. In
this way then, `marlin` allows users to run anything from a simple
Gaussian dispersal kernel up to a system governed by species that
passively diffuse around a dynamic thermal range.

We briefly the general CTMC method here, but see @thorson2021a for a
detailed description of the method. Under this framework, movement from
each patch to each other patch in the system in a given timestep *t* for
life stage *a* of species *s* is defined by a movement matrix
$\boldsymbol{M}_{t,s,a}$. $\boldsymbol{M}_{t,s,a}$ is calculated as the
matrix exponent of the sum of the diffusion and taxis matrices scaled by
the width of the time step that the model is running under.

$$
\boldsymbol{M}_{t,s,a} = e^{\Delta t(\boldsymbol{D}_{t,s,a} + \boldsymbol{X}_{t,s,a})} 
$$ {#eq-movement}

Where $\boldsymbol{D}$ is the diffusion matrix and $\boldsymbol{X}$ is
the taxis matrix.The diffusion matrix is a function of a user specified
diffusion rate, see @thorson2021a.

We slightly modify the taxis matrix, denoted as $\tau$ in the following
manner

$$
\tau_{p1,p2,t,s,a} = \begin{cases}
      = (1 + \lambda d_{s,f})e^{h(p2,t,s,a) - h(p1,t,s,a)} & \text{if p2 and p1 are adjacent}\\
     = -\sum_{p' \neq p1} (1 + \lambda d_{s,f})e^{h(p2,t,s,a) - h(p1,t,s,a)} & \text{if p1 = p2}\\
     = 0 & \text{otherwise.}
  \end{cases}
$$ {#eq-taxis}

The $\lambda$ term is a scalar in which the magnitude of the habitat
gradient between patches can be scaled relative to the diffusion
parameter $d$. When the diffusion and taxis matrices are estimated
jointly from actual data in the manner of @thorson2021a, $\lambda$
should be set to zero. However, when users are manually creating both
the diffusion and taxis matrices, the relative strength of diffusion to
taxis must be defined by the user. Setting higher values of $\lambda$
allows users to simulate a scenario where taxis is much more significant
than diffusion.

For both the diffusion and taxis matrix, we allow for the inclusion of
physical barriers to movement (i.e. land). Pairs of patches that are
adjacent but in which one or both patches are a barrier to movement are
set as non-adjacent. The CTMC model then produces movement dynamics that
move around barriers rather than over them. This is different than
setting habitat in a patch to zero; an animal will preferentially move
towards a patch with zero habitat all else being equal, but can move
over a zero habitat patch towards another better patch.

While this CTMC approach to movement simulation is to date not commonly
seen in the marine modeling literature, it has numerous advantages that
warrant its seeming complexity. First, when only diffusion is present,
the model will generalize to the familiar dynamics of a Gaussian
dispersal kernel. However, the time component *t* in @eq-movement allows
the movement model to be run on arbitrarily fine time steps. Second, the
*taxis* model allow for more active habitat choices by species, allowing
us to simulate both preferences of species in space and time
efficiently. Lastly, this form has the advantage that the parameters are
directly estimable from real data. So, if provided with say spatial
abundance data and a tagging study, users can estimate the movement
parameters in the same manner as @thorson2021a, and then pass their
estimated parameters to `marlin` for simulation.

One challenge is that the coefficients of the model matrices inside
@eq-movement can be hard to interpret. To facilitate this, `marlin`
allows users to specify a maximum number of cells that adults or
recruits diffuse in a year. The model will then tune the diffusion
parameter and habitat gradients to achieve this desired level of
movement.

#### Population Growth

@@

where *BH* is the Beverton-Holt recruitment function, *ssb* is
spawning-stock-biomass. $l_a$ is the mean length at at age, $l_{sel}$ is
the length at which on average 50% of individuals are selected by the
fishery, and $\delta_{sel}$ are the additional units of length at which
on average 95% of fish are selected by the fishery.

@@

Where *u* is a log-normally distributed term allowing for a constant
coefficient of variation in the length at age relationship

$$
u_s \sim N(0,\sigma_s)
$$ {#eq-sigma}

@@

$$
  ssb_{t,p,s} = \sum_{a=1}^{N_a}w_{a,s,t}^{\gamma_s} mat_{t,a} n_{t,p,s,a}
$$ {#eq-ssb}

@@

Recruitment follows Beverton-Holt dynamics. We do however allow for five
variants in the timing of density dependence:

@@

where *hab* is a vector of relative recruitment habitat quality in each
patch.

$$
hab_p = r0_p / \sum_p^P{r0_p}
$$ {#eq-hab}

@@

where **d^l^** is the recruitment movement matrix

@@

@@

### Case Studies

We include two examples demonstrating how `marlin` can be used to
support marine spatial planning. In the first, we show how `marlin` can
be used to compare the total and distributional impacts of marine
protected areas design in a heavily fished coral reef ecosystem. In the
second, we demonstrate how `marlin` can be used to assess components of
climate resilience of alternative MPA design strategies.
